Information carried by thalamocortical axons enables self-organization of the barrel field

### 2019-11-19 paper/briefpaper/brief.tex
Some work reducing down the text in the paper

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopy $|$ Center-point $|$ Pattern $|$}

\begin{abstract}
Understanding how genetic information and spontaneous pattern formation
together give rise to the whisker barrels is an important challenge for
developmental neurobiology. The barrel patterning that coalesces in layer IV
has been described formally as a Voronoi tesselation, which is indicative of
an underlying developmental process based on local competitive
interactions. If pattern formation relies only on such intrinsic
self-organising processes, how can the somatotopic order at the periphery be
maintained in the projections from brainstem to thalamus and from thalamus to
cortex? To investigate, we began by extending a computational model of
neocortical arealization by Karbowski \& Ermentrout (2004; J. Comp. Neurosci.)
to generate fields on a two-dimensional cortical sheet. Central to this model
is an assumption that axonal branching and synaptogenesis are coupled, and
that thalamic projections compete for cortical territory under the influence
of molecular signalling gradients. We found that these processes alone are
incapable of generating patterns of cortical fields that are more intricate
than the patterns already present in molecular guidance signals. However, a
full barrel-field pattern emerges under the guidance of as few as four
signalling gradients when the model is extended to introduce competition
between the branches of axons that originate from different thalamic
barreloids. We show that the key requirements for the emergence of realistic
(Voronoi tesselation) patterning are i) at each cortical location
thalamocortical projections compete for a limited number of available
connections, ii) at each location the axonal branching rate of each
thalamocortical projection is reduced by the total branching density of the
other projections, and iii) the density of axon branching for each
thalamocortical projection is conserved. We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures. We show that this minimal model of
barrel self-organisation provides a parsimonious account of data from a range
of classic studies in which intrinsic and extrinsic factors have been
experimentally manipulated.
\end{abstract}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop. Whether or not patterning
governs information processing, models of the development of information
processing circuits can be validated in terms of their ability to account for
biological patterning (\citealp{purves_iterated_1992}
%; see also \citealp{wilson_what_2015,bednar_cortical_2016}
). One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex \cite{woolsey_structural_1970}.
%
%(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
%see \citealp{fox_barrel_2008}).
%
Understanding how this particular
pattern forms might shed light on the interplay of genetic and
extrinsic factors that shape cortical circuits.
%, and on a wealth of
%anatomical, physiological, and behavioural data obtained using this
%model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
rodent primary somatosensory cortex as dense clusters of thalamocortical
axons, which are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical
surface the barrels constitute a map of the arrangement of the whiskers on the
face
%(see \citealp{yamakado_subdivision_1979})
, such that the cells of adjacent barrels respond most strongly and most
quickly to deflection of adjacent whiskers
\citep{armstrong-james_flow_1992}. The barrels reflect subcortical whisker
maps comprising cellular aggregates known as barreloids in the thalamus and
barrelettes in the brainstem.
%\citep{killackey_pattern_1980}.

Each whisker and its corresponding barrelette, barreloid, and barrel can be
identified by its position on a two-dimensional grid
\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976,haidarliu_size_2001}. The
barrels are normally up to twice as long in the axis aligned to the grid
‘rows’ compared to the ‘arcs’, and the rows are visible from day 2
\citep{rebsam_refinement_2002} whereas the arcs become distinct by day 3
\citep{erzurumlu_development_2012,rebsam_refinement_2002}. In metabotropic
glutamate receptor 5 (mGlu5) knockouts the barrel rows develop but the arcs do
not (\citealp{hannan_plc-1_2001} see also \citealp{fox_barrel_2008}), thus
distinct processes likely guide development of the barrels with respect to the
orthogonal grid axes.

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
(\citealp{anton-bolanos_prenatal_2019}, see also
\citealp{anton-bolanos_developmental_2018}). Patterning also depends on a
complex network of axon guidance processes, which includes interactions
between guidance molecules like ephrin-A5 and A7
\citep{miller_epha7-ephrin-a5_2006} and between adhesion molecules like
cadherin-6 and 8 \citep{greig_molecular_2013,bishop_regulation_2000}, and the
binding of p75 to neurotropins
(\citealp{shimogori_fibroblast_2005,bishop_distinct_2002}; see also
\citealp{dye_lifespan_2011,dye_lifespan_2011-1}). This network is orchestrated
by interactions between morphogens Fgf8 and Fgf17 and transcription factors
Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate
\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.

The boundaries of the barrels form polygonal shapes that together create a
Voronoi tesselation across the cortical sheet, wherein all points enclosed by
a given boundary are closer to its center than to the center of any other
boundary \citep{senft_mouse_1991}. This patterning suggests that the barrel
topology may be preserved in the initial projection of the thalamocortical
axons into the cortex, and that a barrel forms by the lateral branching of the
axons from their initial centres of mass, which ceases upon contact with the
axons branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the initial centre
points, which would itself be determined by guidance cues. However, the
assumption of pre-arranged centres is difficult to resolve with the
observation that thalamocortical axons arrive in the cortical plate as an
undifferentiated bundle prior to barreloid formation
(\citealp{agmon_organized_1993}; see \citealp{erzurumlu_development_2012} for
a review).

@@

To this end, the aim of the present study is to establish whether it is
possible for the observed arrangement of barrels to self-organise in a system
governed by reaction-diffusion dynamics, under the guidance of molecular
signalling gradients, and in the absence of pre-defined centre-points. Our
approach is to represent in a computational model the minimal set of
biological assumptions that might lead to the emergence of barrel patterning
in simulation, and, if necessary, to add to this minimal set only assumptions
that can be represented by the local exchange of information between adjacent
cells. The emergence of barrel patterning under such conditions would serve as
an existence proof of the plausibility of the hypothesis that barrel
development is a self-organising process.

@@

\cite{karbowski_model_2004} developed what remains, to the best of our
knowledge, the only dynamical model of cortical arealization capable of
generating predictions about the influence of signalling molecules on the
size, shape, and topological organization of emergent cortical fields. This
model describes competition between thalamocortical projections for territory
along a one-dimensional string of cortical cells, e.g., along an
anterior-posterior axis. The equations of the model define how the density of
connections, $c$, and the density of axon branches, $a$, interact at time $t$
at each cortical location $x$, for $N$ thalamocortical projections indexed by
$i$. The Voronoi tesselation used to describe barrel patterning is a
two-dimensional structure, hence we first extended the original model for
simulation on a two-dimensional cortical sheet, and describe it here as such
(i.e., substituting $x$ with the two-dimensional position vector $\mb{x}$ and
substituting terms of the form $\frac{d}{dx}$ with the spatial gradient
operator $\nabla$).

The original model was derived from the assumption that the rates at which
thalamocortical synapses form and thalamocortical axons branch are
reciprocally coupled, resulting in an expression for the change in $c_i$ as a
function of $a_i$, and for the change in $a_i$ as a function of
$c_i$. Synaptogenesis is modelled as

\begin{equation} \label{eq:dc}
\frac{\partial c_i(\mb{x},t)}{\partial t} =-\alpha c_i(\mb{x},t) +\beta  \left(1 - \sum_{i'=1}^{N} c_{i'}(\mb{x}, t)\right)[a_i(\mb{x},t)]^k.
\end{equation}

Accordingly, when the final term evaluates to zero, the density of
thalamocortical connections decays exponentially at a rate given by the decay
constant $\alpha$. This occurs where the density of connections over all
thalamic projections sum to one; interpreted as a state in which all of the
available synaptic receptors within the cortical dendritic trees are
occupied. Otherwise, the connection density is assumed to vary non-linearly
($k>1$) with the axon branching density.

Changes in axon branching density are assumed to depend also on the influence
of guidance molecules and on the density of branches at adjacent cortical
locations:

%
\begin{equation} \label{eq:da}
\frac{\partial a_i(\mb{x},t)}{\partial t} = \nabla\vcdot\left(D \nabla a_i(\mb{x},t)-a_i(\mb{x},t)\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i(\mb{x},t)}{\partial t} + \chi.
\end{equation}

The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of a quantity referred to in models of heat diffusion as the
`flux' (the term enclosed by parentheses). The central idea in heat diffusion
models is that the change in temperature (over time) at a given location is
proportional to the gradient in the temperature (over space) at that
location. Substituting the branching density for heat, axons from a given
thalamic projection are thus assumed to branch laterally across the cortical
sheet at a rate given by the diffusion constant $D$. Lateral branching is
augmented by a measure of the local alignment between the branching density
and the gradient in the concentration of each of $M$ distinct molecular
signals, $\rho(\mb{x})$. The influence of a given signal (indexed by $j$) on a
given thalamic projection (indexed by $i$), is determined by the parameter
$\gamma_{i,j}$. Interactions for which this value is positive encourage axon
branching in the direction in which the concentration of the signal is
increasing, whereas negative interactions encourage branching in the opposite
direction. Hence the axons of a given thalamic projection are assumed to
branch towards either higher or lower concentrations of each signalling
molecule.

The second term on the right of equation 2 represents an assumption that the
axon branching density decreases when the number of connections being made
increases, and vice versa, i.e., that branches are required to form synaptic
connections. Note that we introduce the term $\chi=0$ here, so that we might
subsequently redefine $\chi$ to introduce the competition between thalamic
projections that we will show to be necessary for the emergence of a Voronoi
tesselation (see Results).

Two points are important to emphasise here. First, notice that the only
difference between thalamic projections is the interaction strengths,
$\gamma$. Hence any differences between the size, shape, and location of
cortical fields that reliably emerge from the dynamics of the model must be
due to differences in the choice of these values only. Second, the formulation
of the model is at this point entirely local, in the sense that simulating the
dynamics requires no information to be communicated from cortical locations to
any but the immediately adjacent locations.

@@

The dynamics of the model (equations 1 and 2) were solved on a hexagonal
lattice of grid elements representing the cortical sheet, under the assumption
that axon branches cannot extend beyond the boundary of the sheet, i.e.,
$\mb{J}_i(\mb{x},t) \bigg\rvert_{boundary} = 0$. We developed software tools
for imposing arbitrary boundary shapes, specified in a scalable vector
graphics (.svg) file format, on the lattice. The dynamics were numerically
integrated using fourth-order Runge-Kutta, and using a {\color{red} finite
  difference method} to compute the flux term (see Supplementary Materials).

@@

We first verified that the system given by Eqs.~\ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\citet{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
produce a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

Complex structure can indeed be observed in protein expression for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008} and Lmo4
\citep{cederquist_lmo4_2013,greig_molecular_2013}. These expression patterns
can therefore provide a theoretical mechanism for the guidance of axons into
distinct regions.
%The question of how such structures are generated becomes
%one of of how gene transcriptional interactions and mRNA diffusion and
%movement achieves this (\cmnt{Fixme: find a ref on structure in gene
%  transcription}).
However, complex and stable structure in gene expression data is not commonly
reported; monotonically varying expression patterns, such as those for Pax6,
Emx2 \citep{shimogori_fibroblast_2005,bishop_regulation_2000},
Sp8~\citep{sahara_sp8_2007} and Coup-tf1~\citep{borello_sp8_2014}, which
display graded expression across the entire isocortex are more
typical. Furthermore, the complex structure observed in the whisker barrel
field has not, to our knowledge, been linked to a set of gene expression
patterns matching the scale of the barrels. To investigate how the whisker
barrel structure might form in the presence of only monotonically varying
morphogen gradients, we turned our attention to the competition for
connections built into the model.

@@

There exists an element of competition in the model system specified by
Eqs.~\ref{eq:dc} \& \ref{eq:da}. The term $\left(1 - \sum_{i=1}^{N}
c_i(\mb{x}, t)\right)$ in Eq.~\ref{eq:dc} ensures that once the available
synaptic receptors in a region have been fully occupied, say by
thalamocortical type $i=1$, there is no longer the possibility for type $i=2$
to form any synapses. This competition turns out to be relatively weak,
because it does not strongly affect the branching density in that region. In
the model, it is possible for two or more TC axon types to have roughly equal
branching density in some region, leading to roughly equal numbers of
connections forming for each TC type, up until the `hard' connection limit is
reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:branch_comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i(\mb{x}, t)}{N-1} \sum_{i' \ne i}^{N} a_{i'}(\mb{x}, t).
\end{equation}
%
which leads to a reduction in the rate of increase of branching density for
axon type $i$ in regions where the branching density of \emph{other} axon
types is high.

This introduces a new parameter, $\epsilon$, which governs the strength of the
branching competition. The factor $1/(N-1)$ ensures that the branching
competition strength scales with the number of TC axon types in the system.

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation represents the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the finite resources available in
each axon for building new processes. We divisively normalize the branching
density, dividing it at each computational step by the mean branching density
per unit area. It is then multiplied by the branching density per unit area at
time 0, to account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$ after it
has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:branch_comp} on the
basis of the system state at time $t-{\delta}t$.

@@

\cmnt{The realistic barrel pattern experiment is enhanced by stronger
  branching competition. This section may be required here or elsewhere.}

To increase the strength of the competition, we could take the power
of the branching density:
%
\begin{equation} \label{eq:branch_comp_pow}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{i' \ne i}^{N} (a_{i'})^l
\end{equation}
%
with $l>1$. This competition term can quickly become excessively large, so to
preserve the geometric shape of the function for small $a$, whilst
representing the idea that the amount of branching competition has an upper
bound, we pass the function through a logistic function.  This leads to the
following branching equation:

%
\begin{equation} \label{eq:branch_comp_pow_logistic}
\chi(\mb{x}, t) = - \frac{\epsilon a_i}{N-1} \sum_{i' \ne i}^{N}
\frac{1}{1 + e^{-m (a_{i'})^l}} - \frac{1}{2}
\end{equation}
%
$\epsilon$ represents the maximum value of the competition term for TC type
$i$. The parameters $l$ and $m$ control the steepness and shape of the
function.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.7\textwidth]{../../boundaries/barreloids_haidarliu_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}

>>>>

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{Barrel cortex $|$ Self-organization $|$ Somatotopy $|$ Center-point $|$ Pattern}

\begin{abstract}
Understanding how genetic information and spontaneous pattern formation
together give rise to the whisker barrels is an important challenge for
developmental neurobiology. The barrel patterning that coalesces in layer IV
has been described formally as a Voronoi tesselation, which is indicative of
an underlying developmental process based on local competitive
interactions. If pattern formation relies only on such intrinsic
self-organising processes, how can the somatotopic order at the periphery be
maintained in the projections from brainstem to thalamus and from thalamus to
cortex?
%
To investigate, we began by extending a existing computational model of axon
branching and synaptogenesis in thalamocortical projections under the
influence of molecular signalling gradients.  We found that, although these
processes alone are incapable of generating patterns of cortical fields that
are more intricate than the patterns already present in the signalling
gradients, the addition of competition between the branches of axons that
originate from different thalamic barreloids allowed a full barrel-field
pattern to emerge in the presence of only two gradients.
%
%However, a
%full barrel-field pattern emerges under the guidance of as few as four
%signalling gradients when the model is extended to introduce competition
%between the branches of axons that originate from different thalamic
%barreloids.
%
%We show that the key requirements for the emergence of realistic
%(Voronoi tesselation) patterning are i) at each cortical location
%thalamocortical projections compete for a limited number of available
%connections, ii) at each location the axonal branching rate of each
%thalamocortical projection is reduced by the total branching density of the
%other projections, and iii) the density of axon branching for each
%thalamocortical projection is conserved.
%
We demonstrate how a purely
self-organising system can initialise a pattern which can then be faithfully
replicated in multiple neural structures.
%We show that this minimal model of
%barrel self-organisation provides a parsimonious account of data from a range
%of classic studies in which intrinsic and extrinsic factors have been
%experimentally manipulated.
\end{abstract}

@@

\dropcap{S}patial patterns in neural connectivity provide clues about the
constraints under which brains evolve and develop.
%Whether or not patterning
%governs information processing, models of the development of information
%processing circuits can be validated in terms of their ability to account for
%biological patterning \citep{purves_iterated_1992}.
%%see also \citealp{wilson_what_2015,bednar_cortical_2016}
One of the most
distinctive patterns of neural connectivity is found in the rodent
barrel cortex \cite{woolsey_structural_1970}.
%(\citealp{woolsey_pattern_1948,woolsey_structural_1970,welker_structure_1974},
%see \citealp{fox_barrel_2008}).
Understanding how this particular pattern forms might shed light on the
interplay of genetic and extrinsic factors that shape cortical circuits.
%, and on a wealth of
%anatomical, physiological, and behavioural data obtained using this
%model system.

The barrels are large cortical columns, identifiable from birth in layer 4 of
primary somatosensory cortex as dense clusters of thalamocortical axons, which
are enclosed by borders a few neurons thick from postnatal day 3
\citep{erzurumlu_development_2012}. In the plane tangential to the cortical
surface the barrels constitute a map of the arrangement of the whiskers on the
face,
%(see \citealp{yamakado_subdivision_1979})
such that the cells of adjacent barrels respond most strongly and most quickly
to deflection of adjacent whiskers \citep{armstrong-james_flow_1992}. The
barrels reflect subcortical whisker maps comprising cellular aggregates known
as barreloids in the thalamus \cite{van_der_loos_barreloids_1976} and
barrelettes in the brainstem.
%\citep{killackey_pattern_1980}.

%Each whisker and its corresponding barrelette, barreloid, and barrel can be
%identified by its position on a two-dimensional grid.
%\citep{zucker_coding_1969,killackey_pattern_1980,van_der_loos_barreloids_1976,haidarliu_size_2001}.
%
%The barrels are normally up to twice as long in the axis aligned to the grid
%‘rows’ compared to the ‘arcs’, and the rows are visible from day 2
%\citep{rebsam_refinement_2002} whereas the arcs become distinct by day 3
%\citep{erzurumlu_development_2012,rebsam_refinement_2002}.
%In metabotropic glutamate receptor 5 (mGlu5) knockouts the barrel rows develop
%but the arcs do not (\citealp{hannan_plc-1_2001} see also
%\citealp{fox_barrel_2008}), thus distinct processes likely guide development
%of the barrels with respect to the orthogonal grid axes.

Barrel formation requires afferent input, driven by whisker stimulation and by
thalamic calcium waves that propagate via mGlu5-receptor activation
\citep{anton-bolanos_prenatal_2019}.
%(\citealp{anton-bolanos_prenatal_2019}, see also
%\citealp{anton-bolanos_developmental_2018}).
Patterning also depends on a complex network of axon guidance processes, which
includes interactions between guidance molecules like ephrin-A5 and A7
\citep{miller_epha7-ephrin-a5_2006} and between adhesion molecules like
cadherin-6 and 8 \citep{bishop_regulation_2000}, and the
binding of p75 to neurotropins
\citep{bishop_distinct_2002}.
% or both citations?
%\citep{shimogori_fibroblast_2005,bishop_distinct_2002}.
%
%(\citealp{shimogori_fibroblast_2005,bishop_distinct_2002}
%; see also \citealp{dye_lifespan_2011,dye_lifespan_2011-1}).
This network is orchestrated by interactions between morphogens Fgf8 and Fgf17
and transcription factors Emx2, Pax6, Sp8, and Coup-tf1
\citep{shimogori_fibroblast_2005,bishop_regulation_2000}, which are expressed
in gradients that mark orthogonal axes and can be manipulated to cause the
barrels to stretch, shrink, shift, skew, and even duplicate \cite{assimacopoulos_fibroblast_2012}.
% A longer list of citations:
%\citep{shimogori_fibroblast_2005,assimacopoulos_fibroblast_2012,borello_sp8_2014,sahara_sp8_2007,ypsilanti_transcriptional_2016,sur_patterning_2005}.
%
While complex structure in protein expression can be observed for some genes,
such as the molecules Bhlhb5 \citep{joshi_bhlhb5_2008}, a complex gene
expression pattern with the complexity of the barrel field has not, to our
knowledge been reported and so the mechanism by which the barrels form seems
likely to be one in which the structure forms by self-organization.

The boundaries of the barrels form a Voronoi tesselation across the cortical
sheet \citep{senft_mouse_1991}. This patterning suggests that the barreloid
topology may be preserved in the initial projection of the thalamocortical
axons into the cortex, and that a barrel forms by the lateral branching of the
axons from their initial centres of mass, which ceases upon contact with the
axons branching from adjacent centres. According to this model the barrel
pattern could be entirely determined by the distribution of the initial centre
points, which would itself be determined by guidance cues. However, the
assumption of pre-arranged centres is difficult to resolve with the
observation that thalamocortical axons arrive in the cortical plate as an
undifferentiated bundle prior to barreloid formation within the thalamus
\cite{agmon_organized_1993}. % Plus erzurumli, probably

@@

We seek to establish whether it is possible for the observed arrangement of
barrels to self-organise in a system governed by reaction-diffusion dynamics,
under the guidance of molecular signalling gradients, and in the absence of
pre-defined centre-points.
%
Our approach is to represent in a computational model the minimal set of
biological assumptions that might lead to the emergence of barrel patterning
in simulation, and, if necessary, to add to this minimal set only assumptions
that can be represented by the local exchange of information between adjacent
cells. The emergence of barrel patterning under such conditions would serve as
an existence proof of the plausibility of the hypothesis that barrel
development is a self-organising process.

@@

Karbowski \& Ermentrout \citep{karbowski_model_2004} developed a dynamical
model of cortical arealization capable of generating predictions about the
influence of signalling molecules on the size, shape, and topological
organization of emergent cortical fields. Their model describes competition
between thalamocortical projections for territory along an anterior-posterior
cortical axis. The equations of the model define how the density of
connections, $c(x,t)$, and the density of axon branches, $a(x,t)$, interact at time $t$
at each cortical location $x$, for $N$ thalamocortical projections indexed by
$i$.
%The Voronoi tesselation used to describe barrel patterning is a
%two-dimensional structure, hence we first extended the original model for
%simulation on a two-dimensional cortical sheet, and describe it here as such
%(i.e., substituting $x$ with the two-dimensional position vector $\mb{x}$ and
%substituting terms of the form $\frac{d}{dx}$ with the spatial gradient
%operator $\nabla$).
%
The model was derived from the assumption that the rates at which
thalamocortical synapses form and thalamocortical axons branch are
reciprocally coupled, resulting in expressions for the change in $c_i$ as a
function of $a_i$, and for the change in $a_i$ as a function of
$c_i$. Synaptogenesis is modelled [in two dimensions, so the variables are
$a_i(\mb{x},t)$ \& $c_i(\mb{x},t)$] as
%
\begin{equation} \label{eq:dc}
\frac{\partial c_i}{\partial t} =-\alpha c_i +\beta  \left(1 - \sum_{i'=1}^{N} c_{i'}\right)[a_i]^k.
\end{equation}
%
When the final term evaluates to zero, the density of thalamocortical
connections decays exponentially at a rate given by $\alpha$. This occurs when
the density of connections over all thalamic projections sums to one;
interpreted as a state in which all available cortical synaptic receptors are
occupied. Otherwise, the connection density is assumed to vary non-linearly
($k>1$) with the axon branching density.

Changes in axon branching density are assumed to depend also on the influence
of guidance molecules and on the density of branches at adjacent cortical
locations:
%
\begin{equation} \label{eq:da}
\frac{\partial a_i}{\partial t} = \nabla\vcdot\left(D \nabla a_i-a_i\sum_{j=1}^{M} \gamma_{i,j}\nabla \rho_j(\mb{x})\right) - \frac{\partial c_i}{\partial t} + \chi.
\end{equation}
%
The first term on the right describes the divergence (indicated by
$\nabla\vcdot$) of a quantity referred to as the `flux of axonal branching'
(the term enclosed by parentheses, called $\mb{J}_i(\mb{x},t)$).
%The central idea in heat diffusion
%models is that the change in temperature (over time) at a given location is
%proportional to the gradient in the temperature (over space) at that
%location. Substituting the branching density for heat, axons from a given
The flux consists of lateral diffusion across the cortical sheet at a rate
given by the diffusion constant $D$, and an interaction term, in which
branching is influenced by gradients in the concentration of $M$ distinct
molecular signals, $\rho(\mb{x})$. The influence of a given signal (indexed by
$j$) on a given thalamic projection (indexed by $i$), is determined by the
parameter $\gamma_{i,j}$ which may be positive or negative such that axons of
a given thalamic projection may branch preferentially towards either higher or
lower concentrations of each signalling molecule.
%Interactions for
%which this value is positive encourage axon branching in the direction in
%which the concentration of the signal is increasing, whereas negative
%interactions encourage branching in the opposite direction. Hence the axons of
%a given thalamic projection are assumed to branch towards either higher or
%lower concentrations of each signalling molecule.

The second term on the right of Eq.~\ref{eq:da} represents an assumption that the
axon branching density decreases when the number of connections being made
increases, and vice versa, i.e., that branches are required to form synaptic
connections. Note that we introduce the term $\chi=0$ here, so that we might
subsequently redefine $\chi$ to introduce competition between thalamic
projections.
%that we will show to be necessary for the emergence of a
%somatotopic pattern.

Two points are important to emphasise: First, notice that the only
difference between thalamic projections are the interaction strengths,
$\gamma_{i,j}$. Any differences between the size, shape, and location of
cortical fields that reliably emerge from the dynamics of the model must be
due to differences in the choice of these values only; Second, the formulation
of the model is at this point entirely local, in the sense that simulating the
dynamics requires no information to be communicated from cortical locations to
any but the immediately adjacent locations.

@@

The dynamics of the model (Eqs.~\ref{eq:dc} \& \ref{eq:da}) were solved on a
hexagonal lattice of grid elements representing the cortical sheet, under the
assumption that axon branches cannot extend beyond the boundary of the sheet,
i.e., $\mb{J}_i(\mb{x},t) \big\rvert_{boundary} = 0$. We developed software
tools for imposing arbitrary boundary shapes on the lattice. The dynamics were
numerically integrated using fourth-order Runge-Kutta, and using a
{\color{red} finite difference method} to compute the flux term (see
Supplementary Materials).

@@

We first verified that the system given by Eqs.~\ref{eq:dc} \& \ref{eq:da}
described a two dimensional extension of the axon branching model in
\cite{karbowski_model_2004}. We showed that in the presence of three guidance
gradients varying only in one dimension (along the long axis of an elliptical
domain), five thalamocortical axon types, each interacting uniquely with the
three guidance gradients, would segregate into five separate regions. We
investigated a number of configurations of the guidance gradients, varying
their shape, $\rho(\mb{x})$, and number, $M$, and their effect on a varying
number of thalamocortical projections, $N$. In general, we were not able to
produce a complexity of structure in the resulting patterns of the
connections, $c(\mb{x},t\rightarrow\infty)$, which was not present in the
guidance patterns.

@@

To investigate how the whisker barrel structure might form in the presence of
only monotonically varying morphogen gradients, we turned our attention to the
competition for connections built into the model. There does exist an element
of competition in the original model (Eqs.~\ref{eq:dc} \& \ref{eq:da}). The
term $\left(1 - \sum_{i=1}^{N} c_i\right)$ in Eq.~\ref{eq:dc} ensures that
once the available synaptic receptors in a region have been fully occupied by
one thalamocortical axon type, there is no longer the possibility for other
types to form synapses. This competition turns out to be relatively weak,
because it does not strongly affect the branching density in that region; it
is possible for two or more TC axon types to have roughly equal branching
density in some region, leading to roughly equal numbers of connections
forming for each TC type, up until the `hard' connection limit is reached.

To introduce competition between the \emph{branching} of thalamocortical axon
types, we introduced the new term $\chi$ given by
%
\begin{equation} \label{eq:comp}
\chi(\mb{x}, t) = - \frac{\epsilon  a_i}{N-1} \sum_{i' \ne i}^{N} f(a_{i'}).
\end{equation}
%
with the choice $f(a) = a$, which leads to a reduction in the rate of increase
of branching density for axon type $i$ in regions where the branching density
of \emph{other} axon types is high. The new parameter $\epsilon$ governs the
strength of the branching competition and the factor $1/(N-1)$ ensures that the
strength scales with $N$.

With the addition of the competition term, $\chi$, it is necessary to
introduce a normalisation step, in order to prevent one TC type from
out-competing every other TC type. Normalisation represents the idea that for
a fixed number of afferent thalamocortical axons, the overall number of
branches must have some upper limit; given the finite resources available in
each axon for building new processes. We divisively normalize the branching
density, dividing it at each computational step by the mean branching density
per unit area. It is then multiplied by the branching density per unit area at
time 0, to account for the possibility that there might be more axons afferent
for TC type $1$ cf.~TC type $2$:
%
\begin{equation} \label{eq:norm}
  a_i(\mathbf{x}, t) = \int_\Omega  a_i(\mathbf{x}, 0) d\sigma \; \frac {a'_i(\mathbf{x}, t)} {\int_\Omega
  a'_i(\mathbf{x}, t) d\sigma}
\end{equation}
%
The prime symbols indicate the use of the intermediate value of $a_i$ after it
has been computed according to Eqs.~\ref{eq:dc}--\ref{eq:comp} on the
basis of the system state at time $t-{\delta}t$.

@@

\cmnt{The realistic barrel pattern experiment is enhanced by stronger
  branching competition:}

To increase the strength of the competition, we choose an alternative form for
$f(a)$ in Eq.~\ref{eq:comp} which has a geometric form for small $a$: $f(a) = \frac{1}{1 + e^{-m[a]^l}} - \frac{1}{2}$. The parameters $l$ and $m$ control the steepness and shape of the
function.

@@

\begin{figure*}
\begin{center}
\includegraphics[width=0.5\textwidth]{../../boundaries/barreloids_haidarliu_graph.png}
\end{center}
%\vspace{-0em}
\caption{Barreloid centres taken from \cite{haidarliu_size_2001} are shown as
  black circles. It is assumed that there are two linearly increasing,
  orthogonal signalling gradients (molecules A \& B) in the thalamus (side
  graphs), such that the location of each barreloid within the graded thalamic
  signal field can differentiate the efferent axons, with the result that
  axons from each barreloid interact to varying degrees with cortical
  signalling gradients. Coloured arrows show, for each barreloid, the
  magnitude (and sign) of the interaction that its thalamocortical projection
  will have with the signalling gradients in the cortex. The mappings from
  barreloid location to interaction strength are linear.}
\label{fig:barreloids}
\end{figure*}
